---
title: 'Prebiotic + FMT MiPro Study: Microbiota Analysis'
author: "Kait Al"
date: "2026-03-18"
output: 
  bookdown::html_document2:
    toc: true
    toc_float: true
    collapse: section
    toc_depth: 3
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Introduction

This document summarizes the microbiota analyses for the Prebiotic + FMT MiPro study. 

Citation: Kait F Al, Suyang Jia, Michael Silverman, Gregor Reid, Jeremy P Burton, Seema Parvathy, Prebiotic Modulation of FMT Donor Microbiota Enhances MASLD-Relevant Taxa and Functions in an In Vitro Gut Model, Journal of Applied Microbiology, 2026; [DOI:10.1093/jambio/lxag074](https://doi.org/10.1093/jambio/lxag074).

### Abstract:
Metabolic dysfunction-associated steatotic liver disease (MASLD, formerly non-alcoholic fatty liver disease) is a prevalent and progressive condition closely linked to gut microbiota composition. Fecal microbiota transplantation (FMT) may help restore a health-associated microbiome, but its efficacy is often limited by inconsistent engraftment of beneficial taxa. Prebiotics may selectively support keystone microbes associated with reduced MASLD risk. This study evaluated two prebiotics, inulin and xylooligosaccharides (XOS), for their ability to modulate the microbiota of healthy FMT donors in an in vitro gut model, focusing on enriching beneficial taxa and functions associated with MASLD resilience. Stool from eight clinically qualified FMT donors was cultured anaerobically for 24 hours with or without prebiotics. Microbiota composition was assessed by 16S rRNA gene sequencing and short-chain fatty acid (SCFA) concentrations were measured using nuclear magnetic resonance. Functional potential was inferred using predictive metagenomic analysis. Prebiotic responses were highly donor-specific, yet both inulin and XOS consistently enriched Bifidobacterium and Bacteroides – genera associated with SCFA production and metabolic health. XOS preferentially enriched Lactobacillus and Parabacteroides, while inulin enhanced Holdemanella and Mediterraneibacter. Functional pathways relevant to MASLD pathophysiology were enriched, including carbohydrate metabolism, vitamin biosynthesis, fatty acid metabolism, and tryptophan degradation. Both prebiotics significantly increased acetate levels, while butyrate showed a donor-dependent increasing trend. These findings suggest that prebiotic supplementation can selectively enrich MASLD-relevant microbial taxa and functions in donor-derived FMT material, supporting their potential as adjuvants to enhance the efficacy and disease-specificity of FMT interventions for MASLD.

# 2. Initial Code

## Load required packages

```{r message = FALSE, warning = FALSE}
library(ALDEx2)
library(ARTool)
library(broom)
library(cowplot)
library(dplyr)
library(DT)
library(emmeans)
library(forcats)
library(ggplot2)
library(ggrepel)
library(gridExtra)
library(kableExtra)
library(knitr)
library(lme4)
library(lmerTest)
library(Maaslin2)
library(microbiome)
library(patchwork)
library(phyloseq)
library(plyr)
library(RColorBrewer)
library(reshape2)
library(rlist)
library(sjmisc)
library(stringr)
library(tidyr)
library(zCompositions)
```

## Import the files

```{r}
counts <- read.table("data/github_counts.txt", 
                     header = TRUE, row.names = 1, sep = "\t", check.names = FALSE, 
                     quote = "", stringsAsFactors = FALSE)
tax <- read.table("data/cutadapt_tax_138.2.txt", 
                  header = TRUE, row.names = 1, sep = "\t", check.names = FALSE, 
                  quote = "", stringsAsFactors = FALSE)
meta<-read.table("data/github_metadata.txt", header=T, row.names = 1, sep='\t', comment.char = "")
```

## Calculate alpha diversity on unfiltered counts table

```{r}
# make a phyloseq object
tax<-as.matrix(tax) 
OTU = otu_table(counts, taxa_are_rows = FALSE)
TAX = tax_table(tax)
physeq = phyloseq(OTU, TAX)
```

From phyloseq R package, calculate various alpha diversity measures

```{r, warning=FALSE}
div<- estimate_richness(physeq, split = TRUE, measures = NULL)
```

From microbiome R package, calculate various dominance indices

```{r}
dom<- dominance(physeq)
```

Merge the data

```{r}
div_all <- data.frame(div, dom)

alpha<-merge(div_all, meta, by=0, all=FALSE)
#write.table(alpha, file="data/alpha_diversity.txt", sep='\t', quote=F)
```

## Filter the counts tables for downstream analyses

```{r}
# what are the dimensions of the original counts file?
dim(counts)

# Combine the counts and tax tables into one tidier working file
t.counts<-t(counts)
ct<-merge(t.counts, tax, by="row.names")

rownames(ct)<-ct$Row.names
ct$Row.names <- NULL
ct$Sequence <- NULL

ct<-unite(ct,"tax.vector", Kingdom:Genus, sep=':', remove=TRUE)
ct$Species <- NULL
```

Check to see how many samples contain > 1000 total reads.

```{r}
ct2<-ct[,1:ncol(ct)-1]
i <- (colSums(ct2) <=1000)
d.s <- ct2[, !i]
dim(d.s)
ncol(ct2)-ncol(d.s)
```

The DNA blank negative control sample was removed. Now filter out SVs that are not at least 0.1 % abundance in any sample or have fewer than 100 total reads across all samples.

```{r}
d.freq <- apply(d.s, 2, function(x) {x/sum(x)})

#keep SVs > 0.1% in any sample
d.0 <- d.s[apply(d.freq, 1, max)>0.001,]

# make relative abundance table
d.freq0 <- data.frame(apply(d.0, 2, function(x){x/sum(x)}))

#filter SVs based on a read count cutoff
count = 100
d.2 <- data.frame(d.s[which(apply(d.s, 1, function(x){sum(x)}) > count),])

row_list<-intersect(rownames(d.0), rownames(d.2))

d_filt<-d.s[rownames(d.s) %in% row_list,]

#add taxonomy back in and save the filtered counts file
d_filt$tax.vector = ct$tax.vector[match(rownames(d_filt), rownames(ct))]
#write the file
#write.table(d_filt, file="data/dada2_tax_counts_001filt_138.2_gen.txt", sep='\t', quote=F)
```

This file 'd_filt' will be the one we primarily use downstream.

# 3. Figures

## Figure 1. Microbiota composition of donor stool samples following 24 h growth in vitro.

### 1A. Relative abundance barplot

```{r}
gen<-ddply(d_filt, "tax.vector", numcolwise(sum))
taxa<-gen$tax.vector
row.names(gen)<-gen$tax.vector

split6 <- sapply(strsplit(as.character(taxa), ":"), "[", 6)
split6 <- as.data.frame(split6)
rownames(split6)<-rownames(gen)
gen$tax.vector<-NULL

# relative abundance
gen.f <- apply(gen, 2, function(x) {x/sum(x)})

# order by abundance
y1 <- gen.f[order(rowSums(gen.f), decreasing = TRUE),]
y1 <-y1[,-which(colnames(y1)=="PCR_blank")] #remove PCR_blank sample

#check sample columns sums to 1 (100%)
colSums(y1)

bugnames<-rownames(y1)
```

Group together all the rare taxa that are not greater than a specific abundance cutoff. You can try multiple cutoffs, and use one that ensures the barplot is informative but not too busy. Here we decided on 0.5%.

```{r}
abund <- 0.005
keep.taxa.index = rownames(y1[rowMeans(y1) > abund,])

y3 <- as.data.frame(y1) %>% 
  filter(rownames(.) %in% keep.taxa.index) %>% 
  sjmisc::rotate_df() %>% 
  mutate(remainder= 1- rowSums(.)) %>%
  sjmisc::rotate_df() %>% 
  as.matrix(.)

# barplot in ggplot to get stratified by donor sample
melted<-melt(as.matrix(y3))
melted$Donor <-sub("_.*","",melted$Var2)
melted$Treatment<-sub(".*_","", melted$Var2)
colnames(melted)<-c("Genus", "Sample", "Abundance", "Donor", "Treatment")

# rename the taxa without the whole long string
# separate the taxa string into the different elements
test<-data.frame(str_split_fixed(melted$Genus, ":", 6)) 
test$Genus <- paste(test$X6)
melted$Genus<-test$Genus
melted$Genus <- gsub("\\[|\\]", "", melted$Genus) #remove brackets
melted$Genus[melted$Genus == ""] <- "<0.5% Abundance"

melted$Treatment <- gsub("control","C", melted$Treatment)
melted$Treatment <- gsub("inulin","IN", melted$Treatment)
melted$Treatment <- gsub("xos","XO", melted$Treatment)

y4<-melted #for simplicity
pal1=rep(c("white","#006F6B","#00ADAB","#ACDEE0","#3E783A","#76BB47","#AAD486","#CBC02D","#FCF281","#C36928","#F58123","#F8A96E","#F9DDCA","#851719","#ED2027","#F3786D","#4A2970","#7E6BAD","#B8ACD1","#A71D47","#ED1F6B","#F498B9","#1A3E6F","grey","#2276BD","#53A6DC","#A3D9F6","#8e03ad","#e594f7","#f4d7fa","#B2B3B2","#4d6e50"))


pal2=rlist::list.reverse(pal1)

barplot <- ggplot(data=y4, aes(x=Treatment, y=Abundance, fill=Genus)) +
  geom_bar(aes(), stat="identity", position="stack", color="black", linewidth=0.1) +
  scale_fill_manual(values=pal1) +
  facet_grid(~Donor, scales ="free_x") +
  guides(fill = guide_legend(ncol = 2)) +
  ylab("Relative Abundance") +
  theme_bw() +
  theme(axis.text.x=element_text(size=8), legend.position="right", legend.text = element_text(face = "italic", size = 8),
        legend.key.size = unit(0.5, "cm"))
barplot
```

### 1B. PCA plot

Apply compositional data transformation (CZM method) and CLR transformation.

```{r message = FALSE, warning = FALSE}
d.czm <- cmultRepl(t(y1),  label=0, method="CZM")
```

```{r}
d.clr <- t(apply(d.czm, 1, function(x){log(x) - mean(log(x))}))
```

Perform PCA

```{r, warning = FALSE}
d.pcx <- prcomp(d.clr)

# Define parameters for PCA
sv_positions <- data.frame(d.pcx[["rotation"]])
# Merge on genus table
sv_pos <- merge(sv_positions, split6, by = 0)
# Calculate Euclidean distance
arrow_len <- function(x, y) {
  sqrt((x - 0)^2 + (y - 0)^2)
}
sv_pos_dist <- mapply(arrow_len, sv_pos$PC1, sv_pos$PC2)
sv_pos_dist <- as.data.frame(cbind(sv_pos_dist, sv_pos$Row.names, sv_pos$split6, sv_pos$PC1, sv_pos$PC2))
colnames(sv_pos_dist) <- c("Distance", "SV", "Genus", "PC1", "PC2")
# Filter arrow based on distance
filter <- sv_pos_dist %>% filter(Distance >= 0.15)

d.mvar <- sum(d.pcx$sdev^2)
# Calculate the PC1 and PC2 variance
PC1 <- paste("PC1: ", round(sum(d.pcx$sdev[1]^2)/d.mvar, 3))
PC2 <- paste("PC2: ", round(sum(d.pcx$sdev[2]^2)/d.mvar, 3))

metadata<-tibble::rownames_to_column(meta, "SampleID")
loadings<- data.frame(Variables=rownames(d.pcx$rotation), d.pcx$rotation)
values<-merge(d.pcx$x[,c(1,2)], metadata[,c("SampleID","Donor","Treatment")],
              by.x="row.names", by.y="SampleID", all=F)

values$Treatment <- gsub("control","C", values$Treatment)
values$Treatment <- gsub("inulin","IN", values$Treatment)
values$Treatment <- gsub("xos","XO", values$Treatment)

values$Donor<-factor(values$Donor, levels=c("D1","D2","D3","D4","D5","D6","D7","D8"))
values$Treatment<-factor(values$Treatment, levels=c("C","IN","XO"))
cols <- c("D1" = "mistyrose3","D2" = "paleturquoise4","D3" = "goldenrod1","D4" = "deeppink3","D5" = "#B6E7EA","D6" = "mediumpurple3","D7" = "lightpink","D8" = "forestgreen")

theme_new <- theme_set(theme_bw()) 
pca_p <- ggplot(values, aes(x = PC1, y = PC2)) +
  geom_segment(data = sv_pos, aes(x = 0, y = 0, xend = 25 * PC1, yend = 25 * PC2),
               arrow = arrow(length = unit(1/2, 'picas')),
               color = "grey69", alpha = 0.8, size = 0.15) + # Plot features
  geom_text_repel(data = filter, aes(x = 25 * as.numeric(PC1), y = 25 * as.numeric(PC2)), nudge_x = 0.25, nudge_y = 0.25,
                  segment.color = 'transparent',
                  force_pull = 50,
                  force = 7,
                  label = filter$Genus, color = "grey69", size = 3, fontface = "italic") +
  geom_point(data = values, aes(fill=Donor, colour=Donor), shape=21, stroke=1, size=6) +
  geom_text(data = values, aes(label=Treatment), size=2.5) +
  scale_fill_manual(values = cols %>% sapply(function(x) adjustcolor(x, alpha.f = 0.5))) +  # 25% fill opacity
  scale_colour_manual(values = cols) +
  xlab(paste0("PC1: ", round(100*(d.pcx$sdev[1]^2/sum(d.pcx$sdev^2)),1),"%")) +
  ylab(paste0("PC2: ", round(100*(d.pcx$sdev[2]^2/sum(d.pcx$sdev^2)),1),"%")) +
  theme(legend.position = c(0.05, 0.7))
pca_p
```

### 1C. Aitchison distance comparison

Compute Aitchison distances from the CLR transformed abundances.

```{r}
aitch_dists <- as.matrix(dist(d.clr))
```

Make table of distances between C vs other samples with use of loop through Aitchison distance matrix.

```{r}
# Initialize empty lists to hold distances for each category
pIN <- list() # List for IN distances
pXO <- list() # List for XO distances

# Loop through the distance matrix for IN distances
for (i in 1:nrow(aitch_dists)) {
  for (j in 1:ncol(aitch_dists)) {
    currentrow <- rownames(aitch_dists)[i]
    currentcol <- colnames(aitch_dists)[j]
    
    currentrow_split <- unlist(strsplit(currentrow, "_"))
    currentcol_split <- unlist(strsplit(currentcol, "_"))
    
    if (length(currentrow_split) >= 2 & length(currentcol_split) >= 2) {
      if (currentrow_split[1] == currentcol_split[1] & 
          currentrow_split[2] == "inulin" & 
          currentcol_split[2] == "control") {
        # Aaitch_dists distance to the list for IN distances
        pIN[[currentrow_split[1]]] <- c(pIN[[currentrow_split[1]]], aitch_dists[i, j])
      }
    }
  }
}

# Loop through the distance matrix for XO distances
for (i in 1:nrow(aitch_dists)) {
  for (j in 1:ncol(aitch_dists)) {
    currentrow <- rownames(aitch_dists)[i]
    currentcol <- colnames(aitch_dists)[j]
    
    currentrow_split <- unlist(strsplit(currentrow, "_"))
    currentcol_split <- unlist(strsplit(currentcol, "_"))
    
    if (length(currentrow_split) >= 2 & length(currentcol_split) >= 2) {
      if (currentrow_split[1] == currentcol_split[1] & 
          currentrow_split[2] == "xos" & 
          currentcol_split[2] == "control") {
        # Aaitch_dists distance to the list for XO distances
        pXO[[currentrow_split[1]]] <- c(pXO[[currentrow_split[1]]], aitch_dists[i, j])
      }
    }
  }
}

# Now combine the results into a table
# Ensure the participants (prefixes) are unique
all_participants <- unique(c(names(pIN), names(pXO)))

# Create a matrix with the participants as rows and "IN", "XO" as columns
pdist_table <- data.frame(matrix(NA, nrow = length(all_participants), ncol = 2))
rownames(pdist_table) <- all_participants
colnames(pdist_table) <- c("IN", "XO")

# Fill the table with the distances for each category
for (participant in all_participants) {
  if (!is.null(pIN[[participant]])) {
    pdist_table[participant, "IN"] <- mean(pIN[[participant]], na.rm = TRUE)
  }
  if (!is.null(pXO[[participant]])) {
    pdist_table[participant, "XO"] <- mean(pXO[[participant]], na.rm = TRUE)
  }
}

datatable(pdist_table, options=list(paging=FALSE), colnames = c("Donor", "C-IN", "C-XOS"))
```

Now make the plot.

```{r}
# Reset row names to be a column (Donor)
pdist_table$Donor <- rownames(pdist_table)

# Ensure the first 2 columns (BL, IN, XO) are numeric
pdist_table[, 1:2] <- lapply(pdist_table[, 1:2], as.numeric)

# Reshape the data into long format using melt
long_pd <- melt(pdist_table, id.vars = "Donor", 
                measure.vars = c("IN", "XO"), 
                variable.name = "Treatment", 
                value.name = "Distance")

aitch_p <- ggplot(long_pd, aes(x=Treatment, y=Distance)) +
  geom_point(aes(fill=Donor, colour=Donor), shape=21, size=5, stroke=1) +
  scale_fill_manual(values = cols %>% sapply(function(x) adjustcolor(x, alpha.f = 0.25))) +  # 25% fill opacity
  scale_colour_manual(values = cols) +  # Opaque stroke matching the fill color
  ylab("Aitchison Distance") +
  scale_x_discrete(labels=c("C-IN","C-XOS")) +
  stat_summary(aes(x = Treatment, y = Distance), fun = median, geom = "point", shape = 95, size = 10, inherit.aes = FALSE) +
  theme_bw() +
  theme(axis.title.x=element_blank(), legend.position="none")
aitch_p

```

### 1D. Shannon's index plot

First lets do some stats to compare alpha diversity metrics pairwise between C, IN, and XOS samples from the same donors.

```{r, warning = FALSE}
# Subset first 17 columns + Donor and Treatment
metric <- colnames(alpha)[c(2,3,5,7:17)]

# Ensure factors
alpha$Treatment <- factor(alpha$Treatment, levels = c("control", "inulin", "xos"))
alpha$Donor <- factor(alpha$Donor)

# Subset first alpha diversity columns + Donor and Treatment
alpha_sub <- alpha %>%
  dplyr::select(Row.names, metric, Donor, Treatment)

# Reshape to long format
alpha_long <- alpha_sub %>%
  pivot_longer(
    cols = all_of(metric),
    names_to = "Metric",
    values_to = "Value"
  )

# Function: ANOVA with post hoc, returning data frame
run_anova_df <- function(df) {
  all_results <- list()
  
  for (metric in unique(df$Metric)) {
    sub_df <- df %>% filter(Metric == metric)
    # Linear model with donor as blocking factor
    model <- lm(Value ~ Treatment + Donor, data = sub_df)
    # ANOVA p-value for Treatment
    anova_tab <- anova(model)
    anova_p <- anova_tab["Treatment", "Pr(>F)"]
    
    # Post hoc with Tukey adjustment
    posthoc <- emmeans(model, pairwise ~ Treatment, adjust = "tukey")
    posthoc_df <- as.data.frame(posthoc$contrasts)
    posthoc_df <- posthoc_df %>%
      dplyr::mutate(Metric = metric, ANOVA_p = anova_p) %>%
      dplyr::rename(Comparison = contrast, PostHoc_p = p.value) %>%
      dplyr::select(Metric, ANOVA_p, Comparison, estimate, SE, df, t.ratio, PostHoc_p)
    
    all_results[[metric]] <- posthoc_df
  }
  bind_rows(all_results)
}

# Run and collect into one data frame
anova_results_df <- run_anova_df(alpha_long)
anova_results_df <- anova_results_df %>%
  mutate(ANOVA_p = signif(ANOVA_p, 3),PostHoc_p = signif(PostHoc_p, 3)) #round to 3 significant digits

datatable(anova_results_df, 
          options = list(scroll= TRUE, scrollY='300px', paging=FALSE),
          caption = htmltools::tags$caption(style = 'caption-side: top; text-align: left;', htmltools::em('Alpha Diversity ANOVA and Tukey Post Hoc Results')))

```

So both prebiotics seem to reduce alpha diversity, only XOS is statistically significant for Shannon's index.

Now lets make the figure.

```{r}
p_shannon<- ggplot(alpha %>% filter(!is.na(Treatment)), aes(x=Treatment, y=Shannon)) +
  geom_point(aes(fill=Donor, colour=Donor), shape=21, size=5, stroke=1) +
  scale_fill_manual(values = cols %>% sapply(function(x) adjustcolor(x, alpha.f = 0.25))) +  # 25% fill opacity
  scale_colour_manual(values = cols) +  # Opaque stroke matching the fill color
  ylab("Shannon's Index") +
  scale_x_discrete(labels=c("C","IN","XO")) +
  stat_summary(aes(x = Treatment, y = Shannon), fun = median, geom = "point", shape = 95, size = 10, inherit.aes = FALSE) +
  annotate("text", x = 3, y = 3.75, label ="*", size=5) +
  theme_bw() +
  theme(axis.title.x=element_blank(), legend.position = "none")
p_shannon
```

### Build the figure panel

```{r fig.cap="Figure 1. Microbiota composition of donor stool samples following 24 h growth in vitro "}
lay<-rbind(c(1,1,1),c(1,1,1),c(2,2,3),c(2,2,4))
grid.arrange(barplot, pca_p, aitch_p, p_shannon, layout_matrix=lay)
```

## Figure 2. Prebiotic treatment significantly altered the microbiota composition.

First we must perform differential taxonomic abundance analysis using ALDEx2. These values will then be used to make the effect size strip plots in Figure 2 A and B.

These results will not be identical each time you repeat these analyses. This is because ALDEx2 uses Monte Carlo sampling to estimate distributions from compositional data, and each run draws slightly different random samples from those estimated distributions, so the resulting effect sizes and test statistics vary slightly.

```{r, warning=FALSE, message=FALSE}

cont<-grep("_control", colnames(d_filt))
inulin<-grep("_inulin", colnames(d_filt))
xo<-grep("_xos", colnames(d_filt))

aldex.in<-d_filt[,c(inulin,cont)]

#Make a vector of conditions. This must be in the same order and the same number as the columns (samples) of the input table (aldex.in)
conds.in<-c(rep("inulin", length(inulin)), rep("cont", length(cont)))

# get the clr values: this is the main ALDEx function for all downstream analyses
#mc.samples=128 is often sufficient
x.in <- aldex.clr(aldex.in, conds.in, mc.samples=128, verbose=TRUE)

#perform t-test (both Welches and Wilcoxon, plus a Benjamini-Hochberg multiple test correction)
x.tt.in <- aldex.ttest(x.in, paired.test=TRUE)

#estimate effect size and the within and between condition values
#include indiv. samples or not
x.effect.in <- aldex.effect(x.in, paired.test=TRUE)

#merge the data
x.all.in <- data.frame(x.tt.in, x.effect.in)

x.effect.in$tax.vector = d_filt$tax.vector[match(rownames(x.effect.in), rownames(d_filt))]
x.effect.in <- tibble::rownames_to_column(x.effect.in, var="SV")

datatable(
  x.effect.in,
  options = list(
    scrollY='300px',
    scrollX= TRUE,
    paging= FALSE
  ),
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left;',
    htmltools::em('ALDEx2 Effect Size Test: Control vs. Inulin at SV level')
  )
)
```

Now repeat, testing the comparison between control and xos treated samples.

```{r, message=FALSE, warning=FALSE}
aldex.xo<-d_filt[,c(xo,cont)]
conds.xo<-c(rep("xo", length(xo)), rep("cont", length(cont)))
x.xo <- aldex.clr(aldex.xo, conds.xo, mc.samples=128, verbose=TRUE)
x.tt.xo <- aldex.ttest(x.xo, paired.test=TRUE)
x.effect.xo <- aldex.effect(x.xo, paired.test=TRUE)
x.all.xo <- data.frame(x.tt.xo, x.effect.xo)

x.effect.xo$tax.vector = d_filt$tax.vector[match(rownames(x.effect.xo), rownames(d_filt))]
x.effect.xo <- tibble::rownames_to_column(x.effect.xo, var="SV")

datatable(
  x.effect.xo,
  options = list(
    scrollY='200px',
    scrollX= TRUE,
    paging= FALSE
  ),
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left;',
    htmltools::em('ALDEx2 Effect Size Test: Control vs. XOS at SV level')
  )
)
```

### 2A. Control vs Inulin (SV level) strip plot

```{r}
x.effect.in <- x.effect.in %>%
  separate(col="tax.vector",
           into=c("Kingdom","Phylum","Class","Order","Family","Genus"),
           sep= ":",
           remove= TRUE)

sv_in<-subset(x.effect.in, !grepl("Methanobacteriota", Phylum))

theme_new<-theme_set(theme_bw())
pos <- position_jitter(height = 0.2)

SV_inulin<-ggplot(sv_in, aes(x=effect, y=Phylum, label=Genus)) + 
  geom_point(aes(colour = cut(effect, c(-Inf, -0.5, 0.5, Inf)), alpha=cut(effect, c(-Inf, -0.5, 0.5, Inf))), 
             size = 3, position = pos) +
  scale_color_manual(name = "enriched",
                     values = c("(-Inf,-0.5]" = "black",
                                "(-0.5,0.5]" = "black",
                                "(0.5, Inf]" = "steelblue1")) +
  scale_alpha_manual(name = "enriched",
                     values = c("(-Inf,-0.5]" = 0.8,
                                "(-0.5,0.5]" = 0.1,
                                "(0.5, Inf]" = 0.8)) +
  xlab("Effect Size") +
  scale_y_discrete(limits=rev) +
  geom_text_repel(aes(label = ifelse(effect > 0.75, Genus, "")), position = position_jitter(height=0.2), size=2.5, max.overlaps = Inf, alpha=0.5) +
  theme(legend.position="none")
SV_inulin
```

### 2B. Control vs XOS (SV level) strip plot

```{r}
x.effect.xo <- x.effect.xo %>%
  separate(col="tax.vector",
           into=c("Kingdom","Phylum","Class","Order","Family","Genus"),
           sep= ":",
           remove= TRUE)

sv_xo<-subset(x.effect.xo, !grepl("Methanobacteriota", Phylum))

theme_new<-theme_set(theme_bw())
pos <- position_jitter(height = 0.2)

SV_xos<-ggplot(sv_in, aes(x=effect, y=Phylum, label=Genus)) + 
  geom_point(aes(colour = cut(effect, c(-Inf, -0.5, 0.5, Inf)), alpha=cut(effect, c(-Inf, -0.5, 0.5, Inf))), 
             size = 3, position = pos) +
  scale_color_manual(name = "enriched",
                     values = c("(-Inf,-0.5]" = "black",
                                "(-0.5,0.5]" = "black",
                                "(0.5, Inf]" = "indianred1")) +
  scale_alpha_manual(name = "enriched",
                     values = c("(-Inf,-0.5]" = 0.8,
                                "(-0.5,0.5]" = 0.1,
                                "(0.5, Inf]" = 0.8)) +
  xlab("Effect Size") +
  scale_y_discrete(limits=rev) +
  geom_text_repel(aes(label = ifelse(effect > 0.75, Genus, "")), position = position_jitter(height=0.2), size=2.5, max.overlaps = Inf, alpha=0.5) +
  theme(legend.position="none")
SV_xos
```

### 2C-K. The CLR-transformed relative abundance of nine keystone genera.

```{r, warning=FALSE}
clrm<-merge(d.clr, meta, by=0, all=FALSE)

# separate the taxa string into the different elements
test<-colnames(clrm)[2:101]
test1<-data.frame(str_split_fixed(test, ":", 6))
# Create a new column based on the 6th element
new_cols <- ifelse(test1$X6 == "NA", paste(test1$X5, test1$X6, sep = "_"), test1$X6)

# Update column names
colnames(clrm)[2:101] <- new_cols
	
# Define the list of y variables
y_variables <- c(
  "Akkermansia",
  "Bacteroides",
  "Bifidobacterium",
  "Coprococcus",
  "Faecalibacterium",
  "Lachnospira",
  "Lactobacillus",
  "Roseburia",
  "Ruminococcus"
)

# Loop through the variables and create plots
plots <- list()

for (y_var in y_variables) {
  plot <- ggplot(clrm, aes_string(x = "Treatment", y = y_var, group = "Donor")) +
    geom_point(aes(fill = Donor, colour = Donor), shape = 21, size = 5, stroke = 1) +
    scale_fill_manual(values = cols %>% sapply(function(x) adjustcolor(x, alpha.f = 0.25))) +
    scale_colour_manual(values = cols) +
    ylab("Relative abundance (CLR)") +
    ggtitle(paste0(y_var, " spp.")) +
    scale_x_discrete(labels = c("C", "IN", "XO")) +
    stat_summary(
      aes_string(x = "Treatment", y = y_var),
      fun = median, geom = "point", shape = 95, size = 10, inherit.aes = FALSE
    ) +
    theme_bw() +
    theme(
      axis.title.x = element_blank(),
      axis.title=element_text(size=8),
      axis.test=element_text(size=8),
      legend.position = "none"
    )

  # Save each plot in a list
  plots[[y_var]] <- plot
}

# Optionally, display all plots in a grid or save them
grid.arrange(grobs = plots, ncol = 3)


```

### Build the figure panel

```{r, warning=FALSE, fig.cap="Figure 2. Prebiotic treatment significantly altered the microbiota composition"}

layout<-rbind(c(1,1,1),c(1,1,1),c(2,2,2),c(2,2,2),c(3,4,5),c(6,7,8),c(9,10,11))

combined_plots <- c(list(SV_inulin, SV_xos), plots)
fig2 <- grid.arrange(grobs = combined_plots, layout_matrix = layout, heights = c(1, 1, 1, 1, 2, 2, 2))
```

## Figure 3. Volcano plots display differentially abundant microbiota functional features in response to inulin (A, B) and XOS (C, D) treatments.

First we must perform differential abundance analysis using ALDEx2 on the inferred functional outputs from picrust2 (the EC numbers and functional pathways). These values will then be used to make the volcano plots in Figure 3.

These results will not be identical each time you repeat these analyses. This is because ALDEx2 uses Monte Carlo sampling to estimate distributions from compositional data, and each run draws slightly different random samples from those estimated distributions, so the resulting effect sizes and test statistics vary slightly.

```{r, warning=FALSE, message=FALSE}
ec <- read.table("data/picrust_ec.txt", sep="\t", quote="", header=T, row.names = 1)
ec <- round(ec, digits=0)

cont<-grep("_control", colnames(ec))
inulin<-grep("_inulin", colnames(ec))
xo<-grep("_xos", colnames(ec))

aldex.ec.in<-ec[,c(inulin,cont)]
conds.in<-c(rep("inulin", length(inulin)), rep("cont", length(cont)))

x.ec.in <- aldex.clr(aldex.ec.in, conds.in, mc.samples=128)

x.tt.ec.in <- aldex.ttest(x.ec.in, paired.test=TRUE)

x.effect.ec.in <- aldex.effect(x.ec.in, paired.test=TRUE)

x.all.ec.in <- data.frame(x.tt.ec.in, x.effect.ec.in)

x.effect.ec.in <- tibble::rownames_to_column(x.effect.ec.in, var="EC number")

datatable(
  x.effect.ec.in,
  options = list(
    scrollY='300px',
    scrollX= TRUE,
    paging= FALSE
  ),
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left;',
    htmltools::em('ALDEx2 Effect Size Test: Control vs. Inulin in functional EC numbers')
  )
)
```

Now repeat, testing the comparison between control and xos treated samples.

```{r, warning=FALSE, message=FALSE}
aldex.ec.xo<-ec[,c(xo,cont)]
conds.xo<-c(rep("xo", length(xo)), rep("cont", length(cont)))

x.ec.xo <- aldex.clr(aldex.ec.xo, conds.xo, mc.samples=128)

x.tt.ec.xo <- aldex.ttest(x.ec.xo, paired.test=TRUE)

x.effect.ec.xo <- aldex.effect(x.ec.xo, paired.test=TRUE)

x.all.ec.xo <- data.frame(x.tt.ec.xo, x.effect.ec.xo)

x.effect.ec.xo <- tibble::rownames_to_column(x.effect.ec.xo, var="EC number")

datatable(
  x.effect.ec.xo,
  options = list(
    scrollY='300px',
    scrollX= TRUE,
    paging= FALSE
  ),
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left;',
    htmltools::em('ALDEx2 Effect Size Test: Control vs. XOS in functional EC numbers')
  )
)
```

Now repeat again for Inulin and XOS for funcitonal pathways.

```{r, warning=FALSE, message=FALSE}
paths <- read.table("data/picrust_paths.txt", sep="\t", quote="", header=T, row.names = 1)
paths <- round(paths, digits=0)

cont<-grep("_control", colnames(paths))
inulin<-grep("_inulin", colnames(paths))
xo<-grep("_xos", colnames(paths))

aldex.paths.in<-paths[,c(inulin,cont)]
conds.in<-c(rep("inulin", length(inulin)), rep("cont", length(cont)))

x.paths.in <- aldex.clr(aldex.paths.in, conds.in, mc.samples=128)

x.tt.paths.in <- aldex.ttest(x.paths.in, paired.test=TRUE)

x.effect.paths.in <- aldex.effect(x.paths.in, paired.test=TRUE)

x.all.paths.in <- data.frame(x.tt.paths.in, x.effect.paths.in)

x.effect.paths.in <- tibble::rownames_to_column(x.effect.paths.in, var="Pathway")

datatable(
  x.effect.paths.in,
  options = list(
    scrollY='300px',
    scrollX= TRUE,
    paging= FALSE
  ),
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left;',
    htmltools::em('ALDEx2 Effect Size Test: Control vs. Inulin in functional pathways')
  )
)
```

```{r, warning=FALSE, message=FALSE}
aldex.paths.xo<-paths[,c(xo,cont)]
conds.xo<-c(rep("xo", length(xo)), rep("cont", length(cont)))

x.paths.xo <- aldex.clr(aldex.paths.xo, conds.xo, mc.samples=128)

x.tt.paths.xo <- aldex.ttest(x.paths.xo, paired.test=TRUE)

x.effect.paths.xo <- aldex.effect(x.paths.xo, paired.test=TRUE)

x.all.paths.xo <- data.frame(x.tt.paths.xo, x.effect.paths.xo)

x.effect.paths.xo <- tibble::rownames_to_column(x.effect.paths.xo, var="EC number")

datatable(
  x.effect.paths.xo,
  options = list(
    scrollY='300px',
    scrollX= TRUE,
    paging= FALSE
  ),
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left;',
    htmltools::em('ALDEx2 Effect Size Test: Control vs. XOS in functional pathways')
  )
)
```

### 3A. Control vs Inulin (EC number) volcano plot

```{r}
#colours based on significance!
# add a column
x.all.ec.in$diffexpressed <- "NO"
# if log2Foldchange > 0.5 and pvalue < 0.05, set as "UP" 
x.all.ec.in$diffexpressed[x.all.ec.in$effect > 0.5] <- "UP"
# if log2Foldchange < -0.5 and pvalue < 0.05, set as "DOWN"
x.all.ec.in$diffexpressed[x.all.ec.in$effect < -0.5] <- "DOWN"

#labels based on significance!
x.all.ec.in$delabel <- NA
#x.all.ec.in$delabel[x.all.ec.in$diffexpressed != "NO"] <- rownames(x.all.ec.in)[x.all.ec.in$diffexpressed != "NO"]
x.all.ec.in$delabel[x.all.ec.in$effect > 0.5 | x.all.ec.in$effect < -2] <- rownames(x.all.ec.in)[x.all.ec.in$effect > 0.5 | x.all.ec.in$effect < -2]

p_ec_in <- ggplot(data=x.all.ec.in, aes(x=effect, y=-log10(wi.ep), col=diffexpressed)) +
  geom_point(aes(alpha = ifelse(diffexpressed == "NO", 0.5, 1))) +
  theme_bw() +
  scale_color_manual(values=c("black", "black", "steelblue1")) +
  theme(legend.position="none") +
  labs(x="Effect Size", y="Wilcoxon P (-log10)", title="A) EC numbers: Inulin")
p_ec_in
```

### 3B. Control vs XOS (EC number) volcano plot

```{r}
#colours based on significance!
# add a column
x.all.ec.xo$diffexpressed <- "NO"
# if log2Foldchange > 0.5 and pvalue < 0.05, set as "UP" 
x.all.ec.xo$diffexpressed[x.all.ec.xo$effect > 0.5] <- "UP"
# if log2Foldchange < -0.5 and pvalue < 0.05, set as "DOWN"
x.all.ec.xo$diffexpressed[x.all.ec.xo$effect < -0.5] <- "DOWN"

#labels based on significance!
x.all.ec.xo$delabel <- NA
#x.all.ec.xo$delabel[x.all.ec.xo$diffexpressed != "NO"] <- rownames(x.all.ec.xo)[x.all.ec.xo$diffexpressed != "NO"]
x.all.ec.xo$delabel[x.all.ec.xo$effect > 0.5 | x.all.ec.xo$effect < -2] <- rownames(x.all.ec.xo)[x.all.ec.xo$effect > 0.5 | x.all.ec.xo$effect < -2]

p_ec_xo <- ggplot(data=x.all.ec.xo, aes(x=effect, y=-log10(wi.ep), col=diffexpressed)) +
  geom_point(aes(alpha = ifelse(diffexpressed == "NO", 0.5, 1))) +
  theme_bw() +
  scale_color_manual(values=c("black", "black", "indianred1")) +
  theme(legend.position="none") +
  labs(x="Effect Size", y="Wilcoxon P (-log10)", title="B) EC numbers: XOS")
p_ec_xo
```

### 3C. Control vs Inulin (functional pathway) volcano plot

```{r}
#colours based on significance!
# add a column
x.all.paths.in$diffexpressed <- "NO"
# if log2Foldchange > 0.5 and pvalue < 0.05, set as "UP" 
x.all.paths.in$diffexpressed[x.all.paths.in$effect > 0.5] <- "UP"
# if log2Foldchange < -0.5 and pvalue < 0.05, set as "DOWN"
x.all.paths.in$diffexpressed[x.all.paths.in$effect < -0.5] <- "DOWN"

#labels based on significance!
x.all.paths.in$delabel <- NA
#x.all.paths.in$delabel[x.all.paths.in$diffexpressed != "NO"] <- rownames(x.all.paths.in)[x.all.paths.in$diffexpressed != "NO"]
x.all.paths.in$delabel[x.all.paths.in$effect > 0.5 | x.all.paths.in$effect < -2] <- rownames(x.all.paths.in)[x.all.paths.in$effect > 0.5 | x.all.paths.in$effect < -2]

p_paths_in <- ggplot(data=x.all.paths.in, aes(x=effect, y=-log10(wi.ep), col=diffexpressed)) +
  geom_point(aes(alpha = ifelse(diffexpressed == "NO", 0.5, 1))) +
  theme_bw() +
  scale_color_manual(values=c("black", "black", "steelblue1")) +
  theme(legend.position="none") +
  labs(x="Effect Size", y="Wilcoxon P (-log10)", title="C) Pathways: Inulin")
p_paths_in
```

### 3D. Control vs XOS (functional pathway) volcano plot

```{r}
#colours based on significance!
# add a column
x.all.paths.xo$diffexpressed <- "NO"
# if log2Foldchange > 0.5 and pvalue < 0.05, set as "UP" 
x.all.paths.xo$diffexpressed[x.all.paths.xo$effect > 0.5] <- "UP"
# if log2Foldchange < -0.5 and pvalue < 0.05, set as "DOWN"
x.all.paths.xo$diffexpressed[x.all.paths.xo$effect < -0.5] <- "DOWN"

#labels based on significance!
x.all.paths.xo$delabel <- NA
#x.all.paths.xo$delabel[x.all.paths.xo$diffexpressed != "NO"] <- rownames(x.all.paths.xo)[x.all.paths.xo$diffexpressed != "NO"]
x.all.paths.xo$delabel[x.all.paths.xo$effect > 0.5 | x.all.paths.xo$effect < -2] <- rownames(x.all.paths.xo)[x.all.paths.xo$effect > 0.5 | x.all.paths.xo$effect < -2]

p_paths_xo <- ggplot(data=x.all.paths.xo, aes(x=effect, y=-log10(wi.ep), col=diffexpressed)) +
  geom_point(aes(alpha = ifelse(diffexpressed == "NO", 0.5, 1))) +
  theme_bw() +
  scale_color_manual(values=c("black", "black", "indianred1")) +
  theme(legend.position="none") +
  labs(x="Effect Size", y="Wilcoxon P (-log10)", title="D) Pathways: XOS")
p_paths_xo
```

### Build the figure panel

```{r, warning=FALSE, fig.cap="Figure 3. Volcano plots display differentially abundant microbiota functional features in response to inulin (A, B) and XOS (C, D) treatments"}
grid.arrange(p_ec_in, p_ec_xo, p_paths_in, p_paths_xo, nrow=2)
```

## Figure 4. Prebiotic exposure significantly increases acetate concentration

### 4A. Acetate
```{r, warning=FALSE}
meta_scfa <- filter(meta, Treatment != "NA")
meta_scfa <- meta_scfa[,c(9,10,14,15,18)]

a <- ggplot(meta_scfa, aes(x=Treatment, y=Acetate)) +
  geom_point(aes(fill=Donor, colour=Donor), shape=21, size=5, stroke=1) +
  scale_fill_manual(values = cols %>% sapply(function(x) adjustcolor(x, alpha.f = 0.25))) +
  scale_colour_manual(values = cols) +
  ylab("Acetate mM") +
  scale_x_discrete(labels=c("C","IN","XO")) +
  stat_summary(aes(x = Treatment, y = Acetate), fun = median, geom = "point", shape = 95, size = 10, inherit.aes = FALSE) +
  theme_bw() +  # White background
  theme(axis.title.x=element_blank(),legend.position="none")
```

### 4B. Butyrate

```{r, warning=FALSE}
b <- ggplot(meta_scfa, aes(x=Treatment, y=Butyrate)) +
  geom_point(aes(fill=Donor, colour=Donor), shape=21, size=5, stroke=1) +
  scale_fill_manual(values = cols %>% sapply(function(x) adjustcolor(x, alpha.f = 0.25))) +
  scale_colour_manual(values = cols) +
  ylab("Butyrate mM") +
  scale_x_discrete(labels=c("C","IN","XO")) +
  stat_summary(aes(x = Treatment, y = Butyrate), fun = median, geom = "point", shape = 95, size = 10, inherit.aes = FALSE) +
  theme_bw() +
  theme(axis.title.x=element_blank(),legend.position="none")
```

### 4C. Propionate

```{r, warning=FALSE}
p <- ggplot(meta_scfa, aes(x=Treatment, y=Propionate)) +
  geom_point(aes(fill=Donor, colour=Donor), shape=21, size=5, stroke=1) +
  scale_fill_manual(values = cols %>% sapply(function(x) adjustcolor(x, alpha.f = 0.25))) +
  scale_colour_manual(values = cols) +
  ylab("Propionate mM") +
  scale_x_discrete(labels=c("C","IN","XO")) +
  stat_summary(aes(x = Treatment, y = Propionate), fun = median, geom = "point", shape = 95, size = 10, inherit.aes = FALSE) +
  theme_bw() +
  theme(axis.title.x=element_blank(), legend.position="none")

```

### Build the figure panel

```{r, warning=FALSE}
legend <- get_legend(
  p + theme(legend.position = "right"))

plots <- plot_grid(a, b, p, nrow = 1, align = "hv")
final <- plot_grid(plots, legend, rel_widths = c(1, 0.1))

final
```



::: {.tocify-extend-page data-unique="tocify-extend-page" style="height: 598px;"}
:::
